Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-02-03 Multiresolution_framework/ltx/manuscript.tex
edit to the appendix
edits to the appendix

The EM algorithm alternates between maximizing the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\theta'$, in the E-step the lower bound is maximized with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximizing the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\theta,\boldsymbol\theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \theta'}\left[ \cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximizes the expectation of the complete data log-likelihood. 
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}\label{ap:Xiderivation}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in equation~\eqref{eq:MRA-Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	\end{equation}                                                                                                
	We start by expanding the equation \eqref{eq:appXi0}, giving

\begin{align}\label{eq:Xi0derivation0} 
 \mathbf M_{t+1}&=\mathbf E_{\theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1})^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\mathbf x_t\hat{\mathbf x}_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\mathbf x_{t+1}^\top\right]+\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\right]  \nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top+\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top.
\end{align}
Rearranging  equation \eqref{eq:Xi0derivation0} we get
\begin{equation}\label{eq:Xi0derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}+\mathbf {\hat x}_t\mathbf{\hat x}_{t+1}^\top
\end{equation}
substituting \ref{eq:Xi0derivation1} into \ref{eq:appXi0}, noting the expectation commutes with the sum we get
\begin{equation}
 \boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right)
\end{equation}   
An expression for $\boldsymbol\Xi_1$ in terms of the smoothed states can be also found in a similar way by expanding the covariance matrix,
\begin{equation}\label{eq:appXi1}
 \boldsymbol\Xi_1=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t}\mathbf x_{t}^\top\right],
\end{equation}
therefore we can write
\begin{equation}\label{eq:Xi1derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]=\mathbf P_{t}+\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top,
\end{equation}
substituting \ref{eq:Xi1derivation1} into \ref{eq:appXi1} we get
\begin{equation}
 \boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf P_{t}+\mathbf{\hat x}_{t}\mathbf{\hat x}_{t}^\top\right).
\end{equation}
\newpage
\bibliographystyle{./elsarticle/model2-names} 
\bibliography{MRAIDE}

>>>>

The EM algorithm alternates between maximizing the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\theta'$, in the E-step the lower bound is maximized with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximizing the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\theta,\boldsymbol\theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \theta'}\left[ \cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximizes the expectation of the complete data log-likelihood. 
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}\label{ap:Xiderivation}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in equation~\eqref{eq:MRA-Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	\end{equation}                                                                                                
	We start by computing the cross-covariance matrix as

\begin{align}\label{eq:Xi0derivation0} 
 \mathbf M_{t+1}&=\mathbf E_{\theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1})^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\mathbf x_t\hat{\mathbf x}_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\mathbf x_{t+1}^\top\right]+\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\right]  \nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top+\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top.
\end{align}
Rearranging  equation \eqref{eq:Xi0derivation0} gives
\begin{equation}\label{eq:Xi0derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}+\mathbf {\hat x}_t\mathbf{\hat x}_{t+1}^\top.
\end{equation}
Substituting \ref{eq:Xi0derivation1} into \ref{eq:appXi0}, noting the expectation commutes with the sum yields
\begin{equation}
 \boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right)
\end{equation}   
An expression for $\boldsymbol\Xi_1$ in terms of the smoothed states can be also found in a similar way by calculating the covariance matrix, $\mathbf P_t$.
%\begin{equation}\label{eq:Xi1derivation1}
% \mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]=\mathbf P_{t}+\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top,
%\end{equation}
%substituting \ref{eq:Xi1derivation1} into \ref{eq:appXi1} we get
%\begin{equation}
 %\boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf P_{t}+\mathbf{\hat x}_{t}\mathbf{\hat x}_{t}^\top\right).
%\end{equation}
\newpage
\bibliographystyle{./elsarticle/model2-names} 
\bibliography{MRAIDE}
